### 2014-10-14 doc/thesis/system.tex
Adding text about the tags of table types

Since subtyping is reflexive and transitive,
we could have ommited the reflexive rules \textsc{S-LITERAL} and
\textsc{S-BASE}, as well as the rule \textsc{S-INT2}.
More precisely, we could have used subsumption rules that define
the reflexiveness and transitiveness of subtyping.
In the case of \textsc{S-INT2}, we would not need this rule because
we could use the transitive subsumption rule to derive that
\[
\dfrac{\senv \vdash 1 \subtype \Integer \;\;\;
       \senv \vdash \Integer \subtype \Number}
      {\senv \vdash 1 \subtype \Number}
\]

@@

@@

Table types have only one syntactical form, but they also have one of
these three tags: \emph{closed}, \emph{open}, and \emph{unique}.
The tag \emph{closed} is the tag of the table types that the type
annotations describe.
The tag \emph{open} is the tag of the table types that we can
add new fields through field assignment.
The tag \emph{unique} is the tag of the table types that represent
the type of the table constructors.
The syntactical form of table types is $\{ v_{i}, ..., v_{n} \}$,
where $i$ varies from 1 to $n$ when $n > 0$,
but $i$ can also be 0 when $n = 0$.

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
       \senv \vdash v_{j} \subtype v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{closed} \subtype \{v_{1}, ..., v_{n}\}_{closed}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\forall i \in 1..m \; \senv \vdash v_{i}' \subtype v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{open} \subtype \{v_{1}', ..., v_{m}', ..., v_{n}'\}_{open|closed}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\forall i \in 1..m \; \senv \vdash v \subtype v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{open} \subtype \{v\}_{open|closed}} 
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE4}\\
\dfrac{\forall i \in 1..m \; \senv \vdash v_{i}' \subtype_{n} v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{unique} \subtype \{v_{1}', ..., v_{m}', ..., v_{n}'\}_{unique|open|closed}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE5}\\
\dfrac{\forall i \in 1..m \; \senv \vdash v \subtype_{n} v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{unique} \subtype \{v\}_{unique|open|closed}} 
\end{array}
\end{array}
\]

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2} \;\;\;
       \senv \vdash t_{2} \subtype t_{1}}
      {\senv \vdash k_{1}:t_{1} \subtype k_{2}:t_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash \Const \; k_{1}:t_{1} \subtype \Const \; k_{2}:t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash k_{1}:t_{1} \subtype \Const \; k_{2}:t_{2}}
\end{array}
\end{array}
\]

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD4}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash k_{1}:t_{1} \subtype_{n} k_{2}:t_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD5}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash \Const \; k_{1}:t_{1} \subtype_{n} \Const \; k_{2}:t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD6}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash k_{1}:t_{1} \subtype_{n} \Const \; k_{2}:t_{2}}
\end{array}
\end{array}
\]

Table types can be ambiguous.
For instance, $\{1:\Number, \Number:\String\}$ is an ambiguous
table type because the type of the value stored by key $1$ can be
$\Number$ or $\String$, as $1 \subtype 1$ and
$1 \subtype \Number$.
Table types can get even more ambiguous when we use the dynamic
type in the type of table keys.
For instance, the table type $\{1:\Number, \Number:\String, \Any:\Boolean\}$
is ambiguous because the type of the value stored by key $1$ can be
$\Number$, or $\String$, or $\Boolean$, as $1 \subtype 1$,
$1 \subtype \Number$, and $1 \lesssim \Any$.
The type of the value stored by a key of type $\Number$,
which is not the literal type $1$, can be $\Number$ or $\Boolean$,
as $\Number \subtype \Number$, and $\Number \lesssim \Any$.
To avoid ambiguous table types we use the following rule to check
whether a table type is well formed:
\[
\forall i \not\exists j \; i \not= j \wedge k_{i} \lesssim k_{j}
\]

@@

>>>>

Since subtyping is reflexive and transitive,
we could have omitted the reflexive rules \textsc{S-LITERAL} and
\textsc{S-BASE}, as well as the rule \textsc{S-INT2}.
More precisely, we could have used subsumption rules that define
the reflexiveness and transitiveness of subtyping.
In the case of \textsc{S-INT2}, we would not need this rule because
we could use the transitive subsumption rule to derive that
\[
\dfrac{\senv \vdash 1 \subtype \Integer \;\;\;
       \senv \vdash \Integer \subtype \Number}
      {\senv \vdash 1 \subtype \Number}
\]

@@

\emph{/* Explain how Typed Lua refine union types */}

@@

Our type system includes table types to represent Lua tables.
The syntactical form of table types is $\{ v_{i}, ..., v_{n} \}_{tag}$.
When $n = 0$, then $i = 0$ and it represents the type of an empty table.
When $n > 0$, then $i$ varies from 1 to $n$, and each $v_{i}$
represents the type of a table field.
Table fields represent maps from keys to values, and we restricted
the type of keys to literal types, base types, and the dynamic type;
but we can use any first-level type to represent the type of values.
We made this restriction to the type of keys because the statistics
that we discussed in Section \ref{sec:statistics} showed that most
of the tables are records, lists, and hashes.
The dynamic type is an option when we need a more loosen table type.

We also use the tags \emph{closed}, \emph{open}, and \emph{unique}
to classify table types, though they have only one syntactical form.
The first tag represents the table types that the type annotations,
interface, and userdata declarations describe.
The second tag represents the table types that we can use field assignment
to add new fields to them.
The third tag represents the table types of table constructors.
We have different subtyping rules for table types according to their
tags.
We will discuss these subtyping rules in the next paragraphs to
explain why we needed different tags.

The following subtyping rules show that \emph{closed} table types
have width subtyping, and depth subtyping on $\Const$ fields:

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
       \senv \vdash v_{j} \subtype v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{closed} \subtype \{v_{1}, ..., v_{n}\}_{closed}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2} \;\;\;
       \senv \vdash t_{2} \subtype t_{1}}
      {\senv \vdash k_{1}:t_{1} \subtype k_{2}:t_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash \Const \; k_{1}:t_{1} \subtype \Const \; k_{2}:t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash k_{1}:t_{1} \subtype \Const \; k_{2}:t_{2}}
\end{array}
\end{array}
\]

First, we had only \emph{closed} table types.
However, programmers often define tables that include optional fields,
and the subtyping rules we introduced do not let programmers to use
the table constructor to initialize table types that include optional
fields.
For instance, consider the table type we introduced in
Chapter 3:
\begin{align*}
\{ & ``firstname":\String, \\
   & ``middlename":\String \cup \Nil,\\
   & ``lastname":\String \}_{closed}
\end{align*}
Following we show a table type that Typed Lua could assign to a
table constructor that does not include the field $``middlename"$,
but such table type is not subtype of the above table type:
\begin{align*}
\{ ``firstname":\String, ``lastname":\String \}_{closed}
\end{align*}

We introduced \emph{open} table types to allow the refinement
of table types, and also to allow programmers to omit optional
fields in the initialization of table types.
To do that, we defined new subtyping rules that allow \emph{open}
table types to omit fields when $\Nil$ is subtype of the of type
of the value of the omitted key.
The following subtyping rules show this behavior:

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\forall i \in 1..m \; \senv \vdash v_{i}' \subtype v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{open} \subtype \{v_{1}', ..., v_{m}', ..., v_{n}'\}_{open|closed}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\forall i \in 1..m \; \senv \vdash v \subtype v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{open} \subtype \{v\}_{open|closed}} 
\end{array}
\end{array}
\]

However, we still had the that
\begin{align*}
\{ & ``firstname":\String, \\
   & ``middlename":\String,\\
   & ``lastname":\String \}_{open} \not \subtype \\
\{ & ``firstname":\String, \\
   & ``middlename":\String \cup \Nil,\\
   & ``lastname":\String \}_{closed}
\end{align*}

We introduced \emph{unique} table types to solve this problem.
We use \emph{unique} table types only to denote the type of the
table constructor.
The following subtyping rules allow us to use table constructors
that initialize optional fields:

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TABLE4}\\
\dfrac{\forall i \in 1..m \; \senv \vdash v_{i}' \subtype_{n} v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{unique} \subtype \{v_{1}', ..., v_{m}', ..., v_{n}'\}_{unique|open|closed}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE5}\\
\dfrac{\forall i \in 1..m \; \senv \vdash v \subtype_{n} v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{unique} \subtype \{v\}_{unique|open|closed}} 
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD4}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash k_{1}:t_{1} \subtype_{n} k_{2}:t_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD5}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash \Const \; k_{1}:t_{1} \subtype_{n} \Const \; k_{2}:t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD6}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash k_{1}:t_{1} \subtype_{n} \Const \; k_{2}:t_{2}}
\end{array}
\end{array}
\]

Notice that we introduced a special subtyping rule to check the
subtyping of \emph{unique} table fields.
The difference is that \emph{unique} table types have width and
depth subtyping on all fields.

Even though \emph{unique} table types solve the problem of using
table constructors to initialize tables that contain optional fields,
we still have the problem when we try to use a local variable that
holds an \emph{open} table to initialize another table that contains
optional fields.

\emph{/* Do we really need closed, open, and unique? */}

\emph{/* Also, can we make these subtyping rules for tables more clear? */}

Table types can be ambiguous.
For instance, $\{1:\Number, \Number:\String\}$ is an ambiguous
table type because the type of the value stored by key $1$ can be
$\Number$ or $\String$, as $1 \subtype 1$ and
$1 \subtype \Number$.
Table types can get even more ambiguous when we use the dynamic
type in the type of table keys.
For instance, the table type $\{1:\Number, \Number:\String, \Any:\Boolean\}$
is ambiguous because the type of the value stored by key $1$ can be
$\Number$, or $\String$, or $\Boolean$, as $1 \subtype 1$,
$1 \subtype \Number$, and $1 \lesssim \Any$.
The type of the value stored by a key of type $\Number$,
which is not the literal type $1$, can be $\Number$ or $\Boolean$,
as $\Number \subtype \Number$, and $\Number \lesssim \Any$.

To avoid ambiguous table types we use the following rule to check
whether a table type is well formed:
\[
\forall i \not\exists j \; i \not= j \wedge k_{i} \lesssim k_{j}
\]

\emph{/* Can we use this well formed rule to allow $k$ to be $t$? */}

@@

\emph{/* Where should we handle the following case? */}
\begin{align*}
\{ & ``info":\Number, \\
   & ``next":\mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\} \cup \Nil \} \subtype \\
& \mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\}
\end{align*}

\emph{/* Should we also have the following subtyping rules? */}
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-AMBERL}\\
\dfrac{\senv \vdash [x \mapsto \mu x.t_{1}]t_{1} \subtype t_{2}}
      {\senv \vdash \mu x.t_{1} \subtype t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-AMBERR}\\
\dfrac{\senv \vdash t_{1} \subtype [x \mapsto \mu x.t_{2}]t_{2}}
      {\senv \vdash t_{1} \subtype \mu x.t_{2}}
\end{array}
\end{array}
\]

\emph{/* Or having the unfolding typing rule is enough? */}
